Community-Conditioned Language Models

### 2021-05-28 paper/paper.tex
discussion and update

Not only is linguistic variation an object of inquiry in its own
right, but it is also important to consider it in the development of
robust, equitable language technology.  Previous work has shown that
by taking demographic information into account, models can achieve
better performance on various standard NLP tasks
\citep{Hovy2015,Yang2017}, 
%
However, most of the previous computational work on linguistic variation 
has considered variation at the level of macro-social categories, such as gender
\citep{Burger2011,Ciot2013,Bamman2014}, age \citep{Nguyen2013}, and
geographic location \citep{Eisenstein2010,Bamman2014a}.
An exception is \citet{DelTredici2017} who use the modified
skip-gram model of \citet{Bamman2014a} to show that lexical semantic
variation occurs across different communities organized around the
same topic.

@@

  %(Videos)
  %Reminds me of Y2K computer bug. A lot of work went into
  %pre-emptively fixing it, and a lot of bugs were found and fixed,
  %some of which would have been catastrophic. Unfortunately, because
  %the work was successful, a lot of people now describe it as a waste
  %of time/money and a big deal made out of nothing.  }
%
Indeed, it is folklore that certain communities exhibit more
linguistic complexity than others.  Here we want to quantify this
phenomenon.  More precisely, we investigate if the degree of
complexity can be captured in a computational model.  For this purpose
define various Community Conditioned Language Models (CCLM for short),
which can be modeled as an estimator $M$ of the degree to which a
given message $m$ is acceptable in a given community $c$:
\(P_M(m \mid c)\).

Second, we want to find out how, and how much, CCLMs are attuned to
the specificity of communities. We test whether 1.\ the linguistic
models are correlated with a model of communities based only on
co-occurrence of users between communities and 2.\ if latent community
classifiers based on the CCLM can accurately identify that a given
message comes from a given community. Additionally, one may
intuitively assume that linguistically poor communities are easier to
recognize: that is the easiest it is to model computationally
the easier it is to identify.\jp{Do we actually do this?}

@@

Our dataset includes messages from \num{510} subreddits, 
sampled\jp{what is the size of the whole population? why 510?} from the set of all forums 
with at least \num{5000} messages per month for each month
of the year 2015.
Each community corpus consist of \num{42000} randomly selected messages from the year 2015.
For each community, we reserve \num{1000} messages each for development and testing,
leaving a total of \num{20.4}M messages for training.

@@

However, we observe that community-wise perplexity is only weakly (if at all) 
correlated with community indiscernibility under the corresponding latent 
LMCC (see \cref{tab:ind-corrs} and, for example, \cref{fig:cclm-lmcc-ppl}).
In other words, the linguistic complexity of a given community
is a poor predictor of how distinctive its language is.
%
\begin{table}
  \small
  \centering
  \input{floats/ind_corrs.tex}
  \caption{ Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$).  }
  \label{tab:ind-corrs}
\end{table}

@@

Thus if \(d(L,S)\) is below \(422.79\)
($\mu_d - 3s_d$), 
we can safely assume that there 
is statistically significant correlation between \(L\) and
\(S\). 
Indeed, this is very clearly the case (\cref{tab:embedding-correlations}),
meaning that the CCLM embedding capture some of the same information as the
social-network embeddings.
We also give a sense of \emph{how} the
correlation is manifested by analysis of the two principal components
of correlation in the linguistic embeddings, \(U_0\) and \(U_1\). To
do so we plot the projection of embeddings on along \(U_0\) and
\(U_1\), together with the corresponding singular values.

@@

(Another idea may be to correlate the cosine similarity between
\((S_i,S_j)\) and \((L_i,L_j)\). But this would yield one number for
each pair of communities \((510 × 509 / 2)\): too much data to deal with,
and it is hard to get a sense of the meaning, let alone statistical significance of the
resulting correlation coefficient.)\jp{Not sure if this discussion has its place here.}

@@

\jp{I propose that most of the following analysis should be spread between the relevant sections, and only have a tiny summary here + future work.}
We find that various communities exhibit widely varying levels of
linguistic complexity \ref{hyp:varying-complexity}, with perplexity
per-word ranging from less than 34 to more than 65 for the best model 
architecture.

With indiscernibility values evenly ranging over the whole possible
range (0 to 1), we find that CCLMs can be used to recognize certain
communities based on latent Bayesian classification
\ref{hyp:LMCC-works}, but not all.  If generic communities exhibit
distinctive linguistic patterns, none of our models is able to discern
them. In contrast, the distinctive patterns of narrow-topic
communities are just as recognizable by all our LSTM and Transformer
models.

We find no statistically significant correlation (\cref{tab:ind-corrs}) between the
perplexity of the
CCLM and the indiscernibility of the corresponding community
\ref{hyp:rich-harder-to-identify}. Because we have relatively few, and
only large communities in our test set, it is difficult to say if this
finding generalizes. Regardless, this kind of correlation would suggest that
community-specific complexity can be at least partially attributed to
using more varied patterns. This is view is supported by the
observation that, with an indiscernibility $> 0.9$, the most complex
communities appear as an even mixture of the kind of language found in
all communities, to our models.

We find that the layer at which the community embedding is taken into
account in a CCLM influences its perplexity \ref{hyp:layer-effect},
albeit weakly. For LSTM models, the perplexity per word, averaged over
messages from all communities, varies between \num{49.60} and \num{50.83}. (with \num{51.99} for the
unconditioned model) The perplexity per word is slightly higher for
$l_c=0$, and has very little difference between the other models, with
$l_c=3$ being the best.  For transformer models, it varies between 
\num{52.28} and \num{78.76}. Here perplexity per-word is clearly not a monotonous
function of $l_c$: the maximum is reached at $l_c=2$. We did not
expect non-monotonicity, and leave studying this effect to future
work.

The pattern of information gain by community is similar across
architectures.  Communities that benefit most from the conditioned
model behave that way for both the LSTM and transformer.  However,
there are some differences.  For example, the subreddit
\emph{relationships} has an information gain of \num{1.018} for the
best CCLM LSTM but only \num{0.998} for the best transformer model.
Why was the LSTM able to take advantage of the community information
for this community, where the transformer was not?  It would be
interesting to investigate these differences in future work, since it
could reveal differences in the kind of linguistic variation the
different model architectures capture.

The representation of communities built by CCLM exhibits a clear
correlation with user co-occurrence patterns
\ref{hyp:extra-linguistic-correlation}. A natural
question that may arise is how much this correlation can be explain by
user-specific linguistic patterns. We plan to control for this variable
in future work, for example by building author-conditioned language
models and testing if community vector embeddings can be determined by
the set of the vector embeddings of its members.

>>>>

% Not only is linguistic variation an object of inquiry in its own
% right, but it is also important to consider it in the development of
% robust, equitable language technology.  Previous work has shown that
% by taking demographic information into account, models can achieve
% better performance on various standard NLP tasks
% \citep{Hovy2015,Yang2017}. \jp{Move this to related work so we can get to the point quicker.}
=
Most of the previous computational work on linguistic variation 
has considered variation at the level of macro-social categories, such as gender
\citep{Burger2011,Ciot2013,Bamman2014}, age \citep{Nguyen2013}, and
geographic location \citep{Eisenstein2010,Bamman2014a}.
An exception is \citet{DelTredici2017} who use the modified
skip-gram model of \citet{Bamman2014a} to show that lexical semantic
variation occurs across different communities organized around the
same topic.

@@

  %(Videos)
  %Reminds me of Y2K computer bug. A lot of work went into
  %pre-emptively fixing it, and a lot of bugs were found and fixed,
  %some of which would have been catastrophic. Unfortunately, because
  %the work was successful, a lot of people now describe it as a waste
  %of time/money and a big deal made out of nothing.  }
%
Indeed, it is folklore that certain communities exhibit more
linguistic complexity than others.  Here we want to quantify this
phenomenon.  More precisely, we investigate if the degree of
complexity can be captured in a computational model.  For this purpose
define various Community Conditioned Language Models (CCLM for short),
which can be modeled as the degree to which a
given message $m$ is acceptable in a given community $c$, by estimating the probability
\(P(m \mid c)\).

Second, we want to find out how, and how much, CCLMs are attuned to
the specificity of communities. We test whether 1.\ the linguistic
models are correlated with a model of communities based only on
co-occurrence of users between communities and 2.\ if latent community
classifiers based on the CCLM can accurately identify that a given
message comes from a given community. Additionally, one may
intuitively assume that linguistically poor communities are easier to
recognize: that is, the easiest it is to model computationally
the easier it is to identify.

@@

Our dataset includes messages from \num{510} subreddits, 
which all subreddits 
with at least \num{5000} messages per month for each month
of the year 2015.
Each community corpus consist of \num{42000} randomly selected messages from the year 2015.
For each community, we reserve \num{1000} messages each for development and testing,
leaving a total of \num{20.4}M messages for training.

@@

However, we observe that community-wise perplexity is only weakly (if at all) 
correlated with community indiscernibility under the corresponding latent 
LMCC (see \cref{tab:ind-corrs} and, for example, \cref{fig:cclm-lmcc-ppl}).
In other words, the linguistic complexity of a given community
is a poor predictor of how distinctive its language is.
%
\begin{table}
  \small
  \centering
  \input{floats/ind_corrs.tex}
  \caption{ (TODO: missing data) Pearson's $r$ correlation coefficient between community
    indiscernibility ($\Ind_j$) and two different predictors: CCLM
    perplexity on messages from community $c_j$ ($\Ppl_{M_j}$), and
    CCLM information gain on messages from $c_j$ ($\IG_{M_j}$).  }
  \label{tab:ind-corrs}
\end{table}

@@

Thus if \(d(L,S)\) is below the mean by several standard deviations, we can safely
assume that there is statistically significant correlation between
\(L\) and \(S\).  This is the case: we observe a difference between 61 and 68 standard deviations
(\cref{tab:embedding-correlations}). Further, it is clear that the CCLM embedding
capture some of the same information as the social-network embeddings.
We also give a sense of \emph{how} the correlation is manifested by
analysis of the two principal components of correlation in the
linguistic embeddings, \(U_0\) and \(U_1\). To do so we plot the
projection of embeddings on along \(U_0\) and \(U_1\), together with
the corresponding singular values.\jp{TODO}

@@

% (Another idea may be to correlate the cosine similarity between
% \((S_i,S_j)\) and \((L_i,L_j)\). But this would yield one number for
% each pair of communities \((510 × 509 / 2)\): too much data to deal with,
% and it is hard to get a sense of the meaning, let alone statistical significance of the
% resulting correlation coefficient.)\jp{Not sure if this discussion has its place here.}

@@

% \jp{I propose that most of the following analysis should be spread between the relevant sections, and only have a tiny summary here + future work.}
To sum up our findings, 
we observed that various communities exhibit widely varying levels of
linguistic complexity \ref{hyp:varying-complexity}, with perplexity
per-word ranging from less than 34 to more than 65 for the best model 
architecture.

With indiscernibility values evenly ranging over the whole possible
range (1 to 510), we find that CCLMs can be used to recognize certain
communities based on latent Bayesian classification
\ref{hyp:LMCC-works}, but not all.  We cannot rule out that generic
communities exhibit distinctive linguistic patterns, but none of our
CCLMs are able to discern them. In contrast, the distinctive patterns
of narrow-topic communities are just as recognizable by all our LSTM
and Transformer models.

We find a mild correlation (\cref{tab:ind-corrs}) between the
perplexity of the CCLM and the indiscernibility of the corresponding
community \ref{hyp:rich-harder-to-identify}, with factors ranging from
0.06 to 0.23. However, The trend is more clear when considering the
correlation with information gain and indiscernibility (correlation
from 0.39 to 0.46). So this means that, when trying to identify a
community, it is not its absolute perplexity that matters, but rather
the difference in perplexity with the baseline.

We find that the layer at which the community embedding is taken into
account in a CCLM influences its perplexity \ref{hyp:layer-effect},\jp{Delete this hypothesis?}
albeit weakly. For LSTM models, the perplexity per word, averaged over
messages from all communities, varies between \num{49.60} and \num{50.83}. (with \num{51.99} for the\jp{update numbers?}
unconditioned model) The perplexity per word is slightly higher for
$l_c=0$, and has very little difference between the other models, with
$l_c=3$ being the best.  For transformer models, it varies between 
\num{52.28} and \num{78.76}. Here perplexity per-word is clearly not a monotonous
function of $l_c$: the maximum is reached at $l_c=2$. We did not
expect non-monotonicity, and leave studying this effect to future
work.

The pattern of information gain by community is similar across
architectures.  Communities that benefit most from the conditioned
model behave that way for both the LSTM and transformer.  However,
there are some differences.  For example, the subreddit
\emph{relationships} has an information gain of \num{1.018} for the
best CCLM LSTM but only \num{0.998} for the best transformer model.
Why was the LSTM able to take advantage of the community information
for this community, where the transformer was not?  It would be
interesting to investigate these differences in future work, since it
could reveal differences in the kind of linguistic variation the
different model architectures capture.\jp{update?}

The representation of communities built by CCLM exhibits a clear
correlation with user co-occurrence patterns
\ref{hyp:extra-linguistic-correlation}. A natural
question that may arise is how much this correlation can be explained by
user-specific linguistic patterns. We plan to control for this variable
in future work, for example by building author-conditioned language
models and testing if community vector embeddings can be determined by
the set of the vector embeddings of its members.
